---
title: "p8105_hw2_xg2489"
output: github_document
---
```{r}
library(tidyverse)
library(readxl)
```

## Question 1
```{r}
subway_df = 
  read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")
           ) |>
  janitor::clean_names() |>
  select(line:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
      )
  )
dim(subway_df)
```

The orginal dataset "NYC_Transit_Subway_Entrance_And_Exit_Data.csv" has 32 variables related to each entrance and exit for each subway station in NYC. The raw data has 1868 rows and 32 columns.The cleaning process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. Column names is standardized  through "janitor::clean_names", key variables are kept and "entry" is converted from a character variable into a logical variable using "case_match". After cleaning the data , This "subway_df" dataset contains 19 variables including line, station name, station latitude & longitude, routes served (from route 1 to route 11), entry, vending, entrance type, and ADA compliance. The size of dataset is 1868 rows and 19 columns.

This data is not tidy because the routes served by a station are spread across 11 separate columns with too many null values and they should be stored as a single variable. Route number should be a variable, as should route. To obtain a tidy dataset we would need to convert `route` variables from wide to long format.

```{r}
#numbers of distinct stations
subway_df |>
  select(line, station_name) |>
  distinct() |>
  nrow()

#numbers of ADA compliant stations
subway_df |>
  filter(ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()

#proportion of station entrances / exits without vending allow entrance
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()

```

There exists 465 distinct stations and 84 of them are ADA compliant stations. The proportion of station entrances / exits without vending allow entrance is 0.377 approximately.


```{r}
subway_route_df =
  subway_df |>
  pivot_longer(
    route1:route11, 
    names_to = "route_num",
    values_to = "route_name")

#distinct stations serve the A train
subway_route_df |> 
  filter(route_name == "A") |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()    

#distinct stations serve the A train and ADA compliant
subway_route_df |> 
  filter(route_name == "A", ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()     

```

There are 60 stations serving the A train and 17 of them are ADA compliant.


## Question 2

```{r}
mr_trash_wheel_df =
  read_excel("202610 Trash Wheel Collection Data.xlsx",
             sheet = "Mr. Trash Wheel",
             range = "A2:N743"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel_name = "Mr. Trash Wheel"
    )

prof_trash_wheel_df =
  read_excel("202610 Trash Wheel Collection Data.xlsx",
             sheet = "Professor Trash Wheel",
             range = "A2:M140"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel_name = "Professor Trash Wheel"
    )


gwynns_trash_wheel_df =
  read_excel("202610 Trash Wheel Collection Data.xlsx",
             sheet = "Gwynnda the Good Wheel of the W",
             range = "A2:L395"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel_name = "Gwynns Falls Trash Wheel"
    )
```

```{r}

trash_wheel_df = 
  full_join(mr_trash_wheel_df, prof_trash_wheel_df, by = "trash_wheel_name") |>
  full_join(gwynns_trash_wheel_df, by = "trash_wheel_name")

```


```{r}
prof_total_weight = prof_trash_wheel_df |>
  summarize(total = sum(weight_tons)) |>
  pull()

gwynnda_june_cigg = gwynns_trash_wheel_df |>
  filter(month == "June", year == 2022) |>
  summarize(total = sum(cigarette_butts)) |>
  pull() 
```

The combined `trash_wheel_df` dataset contains `r nrow(trash_wheel_df)` observations,
with one row per dumpster collected by any of the three Trash Wheels
(Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda). Key variables
include `dumpster`, `month`, `year`, `date`, `weight_tons`,
`volume_cubic_yd`, and counts of specific item types such as
`plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`,
`plastic_bags`, `wrappers`, and `sports_balls`. The `trash_wheel_name`
variable indicates which machine the record belongs to. Professor
Trash Wheel collected a total of `r round(prof_total_weight, 2)` tons
of trash, and Gwynnda collected `r gwynnda_june_cigg` cigarette butts
in June 2022.


## Question 3
```{r}
baseline_df = 
  read_csv("data_mci/MCI_baseline.csv",skip = 1,na = c("","NA",".")) |>
  janitor::clean_names() |>
  filter(is.na(age_at_onset) | age_at_onset >= current_age) |>
  mutate(
    sex = case_match(               
      sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = case_match(
      apoe4,1 ~ "carrier", 0 ~ "non_carrier"   
    )
  )

mci_num = baseline_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

average_age = baseline_df |>
  pull(current_age) |>
  mean() |>
  round(2)

prop_female = baseline_df |>
  filter(sex == "female") |>
  summarise(p = mean(apoe4 == "carrier")) |>
  pull(p)
```


During the import process, the dataset is imported using skip = 1 because the first row contains variable descriptions. Missing values which are encoded as "." in the orginal dataset are transformed to na argument so that R treats them as NA. Records with an onset age earlier than the baseline age are removed as they do not meet the stated inclusion criteria (i.e. no MCI at baseline). The variables "sex" and "apoe4" are transformed from binary numeric codes into factors ("Female"/"Male" and "Non-carrier"/"Carrier") because they are categorical variables. A total of 483 participants were recruited. After excluding 3 participants who did not meet the inclusion criterion, the cleaned dataset contains 480 participants. The variables are `id`, `current_age`, `sex`, `education`, `apoe4` and `age_at_onset`.

Of the 483 recruited participants, 3 were excluded because they have already developed MCI at baseline. Among the remaining 480 participants who met the inclusion criterion, `r mci_num` developed MCI during follow-up. The average baseline age is `r average_age` and the proportion of women in the study who are APOE4 carriers is `r prop_female`.


```{r}
amyloid_df = 
  read_csv("data_mci/mci_amyloid.csv",skip = 1,na = c("","NA",".")) |> 
  #skip first row variable descriptions
  janitor::clean_names() |>
  rename(id = study_id) |>
  #change to "id" for future Combining
  pivot_longer(
    baseline : time_8,
    names_to = "time_point",
    values_to = "amyloid_ratio"
  ) |>
  #reshape into long format
  mutate(
    years = case_match(
      time_point,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8,
    )
    #add time information
  )
```

```{r}
id_baseline = baseline_df |>
  distinct(id) |>
  pull(id)

id_amyloid = amyloid_df |>
  distinct(id) |>
  pull(id)

only_baseline = setdiff(id_baseline, id_amyloid)
only_amyloid = setdiff(id_amyloid, id_baseline)
both = intersect(id_baseline, id_amyloid)


length(only_baseline)  
length(only_amyloid)   
length(both)           
```

The baseline dataset contains 8 unique participants, while the amyloid dataset contains 15 unique participants and there are 472 participants appeared in both datasets. some participants lack amyloid measurements, while others have amyloid data but no baseline demographic record. Since the analysis requires both, only participants present in both datasets are retained.

```{r}
combined_df = amyloid_df |>
  inner_join(baseline_df, by = "id") 
n_distinct(combined_df$id)  
nrow(combined_df)   

write_csv(combined_df, "combined_df.csv")
```
The combined dataset contains 472 participants and 2360 rows, with up to five repeated amyloid measurements per participant. Variables include id, time_point, years, amyloid_ratio, current_age, sex, education, apoe4, and age_at_onset.
