---
title: "p8105_hw2_xg2489"
output: github_document
---
```{r}
library(tidyverse)
library(readxl)
```

## Question 1
```{r}
subway_df = 
  read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(line:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
      )
  )
dim(subway_df)
```

The orginal dataset "NYC_Transit_Subway_Entrance_And_Exit_Data.csv" has 32 variables related to each entrance and exit for each subway station in NYC. The raw data has 1868 rows and 32 columns. Column names is standardized  through "janitor::clean_names", key variables are kept and "entry" is converted from a character variable into a logical variable using "case_match". After cleaning the data , This "subway_df" dataset contains 19 variables including line, station name, station latitude & longitude, routes served (from route 1 to route 11), entry, vending, entrance type, and ADA compliance. The size of dataset is 1868 rows and 19 columns. This data is not tidy because the routes served by a station are spread across 11 separate columns with too many null values and thsy should be stored as a single variable. 

```{r}
#numbers of distinct stations
subway_df |>
  select(line, station_name) |>
  distinct() |>
  nrow()

#numbers of ADA compliant stations
subway_df |>
  filter(ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> nrow()

#proportion of station entrances / exits without vending allow entrance
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()

```

There exists 465 distinct stations and 84 of them are ADA compliant stations. The proportion of station entrances / exits without vending allow entrance is 0.377 approximately.


```{r}
subway_route_df =
  subway_df |>
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11, 
    names_to = "route_num",
    values_to = "route_name")

#distinct stations serve the A train
subway_route_df |> 
  filter(route_name == "A") |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()    

#distinct stations serve the A train and ADA compliant
subway_route_df |> 
  filter(route_name == "A", ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()     

```

There are 60 stations serving the A train and 17 are ADA compliant.


## Question 2

```{r}
mr_trash_wheel_df =
  read_excel("202509_Trash Wheel_Collection_Data.xlsx",
             sheet = "Mr. Trash Wheel",
             range = "A2:N710"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel_name = "Mr. Trash Wheel"
    )

prof_trash_wheel_df =
  read_excel("202509_Trash Wheel_Collection_Data.xlsx",
             sheet = "Professor Trash Wheel",
             range = "A2:M135"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel_name = "Professor Trash Wheel"
    )


gwynns_trash_wheel_df =
  read_excel("202509_Trash Wheel_Collection_Data.xlsx",
             sheet = "Gwynns Falls Trash Wheel",
             range = "A2:L352"
             ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel_name = "Gwynns Falls Trash Wheel"
    )
```

```{r}

trash_wheel_df = 
  full_join(mr_trash_wheel_df, prof_trash_wheel_df, by = "trash_wheel_name") |>
  full_join(gwynns_trash_wheel_df, by = "trash_wheel_name")

```


```{r}
prof_total_weight = prof_trash_wheel_df |>
  summarize(total = sum(weight_tons)) |>
  pull()

gwynnda_june_cigg = gwynns_trash_wheel_df |>
  filter(month == "June", year == 2022) |>
  summarize(total = sum(cigarette_butts)) |>
  pull() 
```

The combined `trash_wheel_df` dataset contains `r nrow(trash_wheel_df)` observations,
with one row per dumpster collected by any of the three Trash Wheels
(Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda). Key variables
include `dumpster`, `month`, `year`, `date`, `weight_tons`,
`volume_cubic_yd`, and counts of specific item types such as
`plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`,
`plastic_bags`, `wrappers`, and `sports_balls`. The `trash_wheel_name`
variable indicates which machine the record belongs to. Professor
Trash Wheel collected a total of `r round(prof_total_weight, 2)` tons
of trash, and Gwynnda collected `r gwynnda_june_cigg` cigarette butts
in June 2022.
