---
title: "p8105_hw2_xg2489"
output: github_document
---
```{r}
library(tidyverse)
```

## Question 1
```{r}
subway_df = 
  read_csv(file  = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(line:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
      )
  )
dim(subway_df)
```

The orginal dataset "NYC_Transit_Subway_Entrance_And_Exit_Data.csv" has 32 variables related to each entrance and exit for each subway station in NYC. The raw data has 1868 rows and 32 columns. Column names is standardized  through "janitor::clean_names", key variables are kept and "entry" is converted from a character variable into a logical variable using "case_match". After cleaning the data , This "subway_df" dataset contains 19 variables including line, station name, station latitude & longitude, routes served (from route 1 to route 11), entry, vending, entrance type, and ADA compliance. The size of dataset is 1868 rows and 19 columns. This data is not tidy because the routes served by a station are spread across 11 separate columns with too many null values and thsy should be stored as a single variable. 

```{r}
#numbers of distinct stations
subway_df |>
  select(line, station_name) |>
  distinct() |>
  nrow()

#numbers of ADA compliant stations
subway_df |>
  filter(ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> nrow()

#proportion of station entrances / exits without vending allow entrance
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()

```

There exists 465 distinct stations and 84 of them are ADA compliant stations. The proportion of station entrances / exits without vending allow entrance is 0.377 approximately.


```{r}
subway_route_df =
  subway_df |>
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11, 
    names_to = "route_num",
    values_to = "route_name")

#distinct stations serve the A train
subway_route_df |> 
  filter(route_name == "A") |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()    

#distinct stations serve the A train and ADA compliant
subway_route_df |> 
  filter(route_name == "A", ada == TRUE) |>
  select(line, station_name) |> 
  distinct() |> 
  nrow()     

```

There are 60 stations serving the A train and 17 are ADA compliant.

## Question 2


